Floral selectivity in mountain bumble bees and a test of the 'target use hypothesis

### 2021-06-09 ms_2/ms_2.Rmd
updated ms2

---
title: "Bumble bees exhibit strong but flexible foraging biases through temporal and elevational gradients of floristic turnover"

author: Douglas B. Sponsler^1^*, Fabrice Requier^2^, Katharina Kallnik^1^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Ingolf Steffan-Dewenter^1^

@@

*Keywords*: optimal foraging, trait matching, pollinator, *Bombus*, alpine

@@

# INTRODUCTION

<!-- Open with a discussion of the problem of foraging efficiently in a dynamic foraging environment, the balance of bias and flexibility, etc. -->

Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies (Plowright and Laverty 1984), or indeed by human investigators (Alem et al. 2016; Loukola et al. 2017), renders almost any flower potentially accessible. Even when a gross mismatch between tongue length and corolla depth makes the nectar of a given flower inaccessible by ordinary means, some bumble bee species will resort readily to biting through the base of the flower to “rob” nectar, incidentally making the nectar available to other prospective foragers as well. Flower handling skills are, moreover, socially transmitted, amplifying the plasticity of individual behavior (Leadbeater and Chittka 2008; Loukola et al. 2017). Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the searching/learning/handling costs associated with exploiting it. At an evolutionary level, the innate biases of bumble bees for certain floral traits likely reflect historic associations between floral traits and floral rewards (Raine and Chittka 2007). These innate biases, however, can be modulated or even nullified by patterns of reward learned during foraging (Kugler 1943; Maharaj et al. 2019), which depend not only on intrinsic floral properties but also ecological modulators, such as the spatial distribution of floral patches or competition with other bumble bees (Ishii 2013; Plowright and Laverty 1984; Sowig 1989).

The topic of floral bias in bumble bees has been approached via both laboratory experimentation (e.g. Kugler 1943; Raine and Chittka 2007) and observational field study (e.g. Heinrich 1976; Dave Goulson, Lye, and Darvill 2008). The former approach enables stronger causal inference with respect to the mechanisms of floral choice, but it suffers from all the usual limitations of biological realism associated with laboratory conditions. Observational field studies, however, tend to be anecdotal in scope, typically describing bumble bee foraging patterns in particular localities, over short time frames, and with limited floristic data to enable a distinction between true foraging bias and the neutral effects of floral abundance. The steep elevation gradients of mountain slopes provide an opportunity for more powerful observational study of bumble foraging because they generate broad gradients of environmental variation while controlling for confounding effects of geographic variation, such as genotypic differentiation within bumble bee species. When bumble bee visitation in such environments is sampled with sufficient spatial and temporal replication and complemented with quantitative floral surveying, true bias can be distinguished from the neutral effects of abundance, and variation in bias can be explored along elevational and temporal dimensions of floristic turnover.

In the present study, we focus on the question of bumble bee foraging bias with respect to floral morphology, and the lability of morphological biases across gradients of floristic variation. At the outset, we emphasize the use of the term “bias” in favor of the term “preference,” since the latter is laden with more assumptions and its inference under field conditions is problematic (Williams 2005). By “bias,” we mean simply an empirical disproportionality between the relative abundance of a given floral taxon or morphological class and its proportional visitation. The exact mechanism of this disproportionality --- whether an arbitrary “preference” on the part of the foraging bee, an unmeasured decoupling of relative abundance and rate of reward, an exogenous effect of competition, etc. --- is beyond the scope of our discussion and probably best studied under laboratory conditions where strict factorial control is possible. We focus instead on the empirical phenomenon of bias, its variation, and its morphological correlations.

The meaning of “morphology,” however, is hardly more clear-cut than that of bias. Bumble bees and flowers vary in an array of interrelated morphological parameters, both within and between species, and the selection of relevant and measurable morphological metrics is nontrivial. For bumble bees, morphological variation consists principally (though not exclusively) of variation in tongue length. Tongue length in bumble bees corresponds functionally to corolla depth in flowers, and there is a well-established (though imperfect) association between tongue length and floral choice (e.g. Inouye 1980). Importantly, though, accessing floral reward involves more than just a match between tongue length and corolla depth. Such matching is entirely bypassed by nectar robbers, but even in the case of “legitimate” foraging, access to reward involves learning to manipulate floral structures that, in some floral morphotypes, are challenging irrespective of tongue length (Laverty 1994). For example, the characteristic Papilionaceous floral morphology of legumes requires a bee to separate tightly appressed keel petals to access anthers and nectaries (Westerkamp 1997), and the narrowly campanulate flowers of many Ericaceous flora pose a significant barrier to large-bodied visitors like bumble bees (Wallace 1977). Rather than considering various floral morphological traits independently, we rely on the synthesis of Kugler (1970), who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice (Kugler 1943), and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes (Neumayer and Paulus 1999; Schneller et al. 2014).

<!-- Formulate this section more as a list of motivating questions; perhaps end with something that ties back into the more general introduction. -->

How do floral morphotypes compare in terms of total bumble bee visitation?

How does morphological bias vary across bumble bee species and tongue length classes?

Species' morphological biases give rise to patterns of diet (dis)similarity

How are the floral morphotypes distributed through elevation and time in our mountain ecosystem?

For ubiquitously distributed bumble bee species, how does morphological bias vary through elevation and time?

First, we compare the overall rate of bumble bee visitation across floral morphotypes. Then, we parse visitation rates by bumble bee species and explore the functional relationship between floral morphotype and bumble bee tongue length. These biases, together with the background effects of relative abundance, generate emergent patterns of diet (dis)similarity, which we present in the form of ordination analysis. Lastly, we analyze the lability of the morphological biases of individual bumble bee species through elevation and time, and compare these patterns of bias across bumble bee species and within floral morphotypes. The results of these analyses demonstrate both the strength of bumble bee floral bias, arising in part from morphological constraints, and the flexibility of bumble bee floral bias that enables them to forage adaptively in the dynamic floral communities of mountain landscapes.

# METHODS

## Field system

<!-- I guess I'll need to tweak the wording of this section so that it is not a verbatim copy of the other manuscript. But it's okay if it recycles a lot of language because it is, after all, the same thing being described. -->

<!-- Describe general floristic and temperature patterns. Describe sampling in more detail. -->

Field sampling was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures in a coniferous forest matrix. We selected 25 study sites (60 x 60 m) on pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management. The sites sampled in this study are the same as those used in a series of previous studies [@Hoiss2012-sn; @Hoiss2012-kr; @Hoiss2015-gv], but the data reported here are otherwise independent.

Sampling consisted of repeated visits (hereafter, “visitation samples”) to each study site at approximately weekly intervals. Samples were only collected on dry days when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on or a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, whereas workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory after Amiet (1996). While floral visitation by male bumble bees was recorded during visitation sampling, we chose to analyze only visitation by queens and workers. During observation, it was noted whether each bee was visibly carrying pollen at the time of its visit, but this provides at best an imperfect distinction between pollen and nectar foragers, and we opted not to differentiate between bees with and without pollen in our analyses. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.1 m^2^. Species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

While per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied markedly,  with 153 in 2010, 212 in 2011, and 539 in 2012. The increased total sampling effort in 2012 was driven in part by the overall scarcity of bumble bees that year. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Appendix X for details of sampling patterns.

## Data analysis

### Data processing

Bumble bee tongue-length data were obtained from the literature values compiled and summarized by Arbetman et al. (2017). Values for *B. gerstaeckeri* and *B. mendax*, which were not included in Arbetman et al. (2017), were obtained from Obeso (1992) and Durieux (2000), respectively. We then binned bumble bee species into three tongue-length classes: short (< 7.5 mm), medium (7.5—10.0 mm), and long (>= 10.0 mm). Short-tongued species included *B. jonellus*, *B. lapidarius*, *B. pratorum*, *B. psithyrus*-group, *B. pyrenaeus*, and *B. terrestris/lucorum*. Medium-tongued species included *B. humilis*, *B. hypnorum*, *B. monticola*, *B. mucidus*, *B. pascuorum*, and *B. wurflenii*. Long-tongued species included *B. gerstaeckeri*, *B. hortorum*, and *B. mendax*.

Floral morphology was summarized using floral morphotypes of Kugler (1970), accessed via the BIOFLOR database (Klotz, Kühn, and Durka 2002). Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed Neumayer and Paulus (1999) and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, so this morphotype was also omitted from all analyses.

Prior to analysis, we aligned each visitation sample with a corresponding floral survey sample. In most cases, floral surveying was conducted jointly with bumble bee observation for each site and date. Occasionally, though, floral surveying could not be completed on the same day as bumble bee observation, resulting in a small temporal offset between visitation samples and floral survey samples. In these cases, visitation samples were paired with the nearest available floral survey data within a maximum distance of 7 days. Visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases, two consecutive visitation samples at a given site shared the same nearest floral survey sample; in these cases, the visitation samples were first pooled, then aligned to their shared nearest floral survey sample. In rare cases, a given floral species was recorded during visitation sampling but missed during floral surveying. These species were retroactively assigned the minimum floral area value of 0.01 m^2.

Visitation data were extremely right-skewed because the majority of available flora received zero visits during a given sampling period. To facilitate model fitting, we created a binary version of visitation rate in which visited flora were assigned to 1 and unvisited flora to 0. We then modeled binary visitation using generalized models with binomial or Bernoulli distribution families. The conditional effects inferred from these models should, therefore, be interpreted as effects on visitation probability rather than visitation frequency.

While our study spanned three years, we chose not to model year effects. Differences in sampling intensity and duration across years, for the reasons described above, make comparisons across years difficult to interpret, and we deemed these comparisons not worth the cost of added model complexity.

<!-- Should I be including year as a varying effect in these models? -->

<!-- State more explicityl how the models should be interpreted withr expect to bias -->

<!-- Report the effects of abundance that I am conditioning on -->

<!-- Perhaps show what happens if we remove the abundance interaction terms -->

<!-- Mor edescriptive fgure captions -->

<!-- Add something about model diagnostics -->

<!-- Consider adding year as a varying effect -->

<!-- Figure 3 order by tongue length class -->

<!-- Describe the differences in models in the Figure2 and 3 captions -->

<!-- Explain filtering of ordination plot -->

<!-- Year should really be included in Model 5, or just use 2012. -->

<!-- Open each results paragraph with a summary sentence. -->

### Model 1: Comparing overall visitation probability across floral morphotypes

We first compared overall visitation probability across floral morphotypes. In this model, visitation for each sample was pooled across bumble bee species and within floral morphotypes prior to binarization. Visitation probability was then modeled using multilevel Bayesian regression with a Bernoulli distribution family (logit link function). *Bumble bee abundance* and *percent flower cover* (per morphotype) were included as a global interaction term, thus conditioning visitation probability on abundance and enabling the inference of bias. *Floral morphotype* was specified as a varying effect term with both a varying intercept and a varying slope with respect to *percent flower cover*, and a varying intercept term was also included for *site*. See Appendix X for the details of model specification and evaluation.

### Model 2: Visitation probability as a function of floral morphotype and bumble bee tongue length

To explore the functional underpinnings of bumble foraging bias, we analyzed visitation probability as a function of the interaction between floral morphology and bumble bee tongue length class. For this analysis, visitation was pooled within floral morphotype but not across bumble bee species. We then specified a Bayesian model of the same form as Model 1, but this time included a varying slope and intercept term for floral morphotype : bumble bee tongue length class and floral morphotype : bumble bee species. In this model, the relationship between floral morphotype and bumble bee tongue length class was the focus of inference, and the bumble bee species effect was added to avoid pseudoreplication arising from the grouping of observation within species. See Appendix X for the details of model specification and evaluation.

### Model 3: Visitation probability as a function of floral morphotype and bumble bee species

To focus on species-level morphotype biases, we removed the tongue length effect from Model 2 and re-fit the model considering only bumble bee species. See Appendix X for the details of model specification and evaluation.

### Model 4: Aggregate diet ordination

To visualize the net diet similarity between bumble bee species, we pooled visitation data across all samples within bumble bee species and plant species, then performed a principal coordinates analysis based on Horn-Morisita distance metric. To test the significance of tongue length class as a predictor of diet similarity, we performed a permutation test.

### Model 5: The variation of morphological bias through elevation and time

To explore the plasticity of floral biases, we first subsetted our data to include only the dominant bumble bee species (B. hortorum, B. pascuorum, B. pratorum, B. soroeensis, B. terrestris/lucorum, and B. wurflenii) and the most visited floral morphotypes (groups 3, 5, 6, and 7). We then constructed a generalized additive model (GAM) in which binary visitation was modeled as a tensor-product smooth of elevation and day-of-year, grouped by floral morphotype. This response was conditioned on linear interaction terms for bumble bee abundance : percent flower cover and percent flower cover : floral morphotype and a random intercept term for bumble bee species : floral morphotype. See Appendix X for the details of model specification and evaluation.

## Software

All analyses were conducted in `R` `r citep(citation())`. Bayesian multilevel models were implemented in Stan (Stan Development Team 2021) via the package `brms` `r citep(citation("brms"))`, and visualizations were constructed with the packages `tidybayes` `r citep(citation("tidybayes"))` and `ggdist` `r citep(citation("ggdist"))`. Ordination analysis and permutation testing were performed with the package `vegan` `r citep(citation("vegan"))` and visualized using the packages `ggplot2` `vegan` `r citep(citation("ggplot2"))` and `ggvegan` (Simpson 2019). GAMs were implemented with the package `mgcv` [@Wood_2017] and visualized with the package `mgcViz` [@Fasiolo2018aa]. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("gdm"))`. Annotated R code is available in the Supplementary Material.

# RESULTS

## Summary of visitation and floral survey data

<!-- Modify this section to describe how many observations/samples were dropped due to the non-alignment of visitation and survey data -->

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species (with *Bombus terrestric/lucorum* and *Bombus psithyrus* species groups lumped into species groups), 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

>>>>

---
title: "Floral selectivity in mountain bumble bees and a test of the 'target use hypothesis'"

author: Douglas B. Sponsler^1^*, Katharina Kallnik^1^, Fabrice Requier^2^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Johanna Sieger^3^, Ingolf Steffan-Dewenter^1^

@@

Alternative title: Mountain bumble bees exhibit strong but flexible foraging biases through temporal and elevational gradients of floristic turnover

The process of resource selection by foraging animals has been a central theme of ecological research since the mid 20th century, particularly in the context of resource partitioning among sympatric species. Bumble bees as model system... Field studies have shown that floral choice in bumble bees is associated with the morphological fittedness between bumble bee tongue length to floral corolla depth. Morphological fittedness, however, is only a soft constraint on resource selection; bumble bees also tune their floral choices to learned patterns of reward, and they exhibit remarkable intelligence in learning the handling skills required to exploit difficult flower morphologies.

Using a dataset consisting of more than 13,000 observations collected along a 1400 m elevation gradient over three consecutive years, we describe how bumble bee floral selection is constrained by morphology yet flexibly tuned to local conditions in elevation and time, with considerable idiosyncrasy in floral selection across species with similar tongue length.  More results...

Significance ...

*Keywords*: foraging, trait matching, pollinator, *Bombus*, alpine

@@

# 1. INTRODUCTION

The process of resource selection by animals has been an enduring theme of ecological research [@Johnson1980-co; @Beyer2010-vr]. Methodological difficulties notwithstanding, the study of animals' resource choices is one of the central tasks of community ecology, where the selection and partitioning of resources are instrumental in community assembly, species coexistence, and ecosystem function [@Chesson2000-aw]. Recently, questions of resource selection in ecological communities have taken on new urgency on account of anthropogenic climate and land use change, which threaten to disrupt existing patterns of species interactions and niche relationships [@Tylianakis2008-fi].

Bumble bees (Hymentoptera: *Bombus* spp.) and their floral hosts provide a diverse yet tractable system in which to explore resource selection. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [@Plowright1984-lc] --- or, indeed, by human investigators [@Alem2016-rn; @Loukola2017-bt] --- renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs of exploiting it. These costs are closely related to the degree of morphological fittedness between bumble bees and their floral hosts, which consists principally in the matching of bumble bee tongue length to floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], leading to a well-documented association between tongue length and floral choice in bumble bee communities [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. This association between tongue length and floral niche is imperfect, though, and there can be considerable overlap in floral resource use between bumble bees of different tongue lengths [e.g. @Ranta1980-ok; @Goulson2008-oe]. Moreover, bumble bee floral choice is also influenced by ecological modulators, such as the spatial distribution of floral patches or competition with other bumble bees [@Ishii2013-sm; @Plowright1984-lc; @Sowig1989-ld].

Understanding resource selection in bumble bee communities is more than just an exercise in ecological theory. Bumble bees are sensitive to climate and land use change [@Goulson2008-oe; @Soroye2020-zb], especially in the mountain ecosystems that host the world's richest bumble bee communities. Warming climates have already caused measurable upslope shifts in mountain bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], which could introduce novel competitive interactions between bumble bee species that experience increased sympatry. At the same time, socioeconomic factors have led to a dramatic decline in traditional alpine grazing management, particularly in the European Alps, resulting in widespread pasture abandonment and reforestation of mountain meadows [@Tappeiner2006-jp; @Tasser2007-qo]. The regression of meadows to forest entails a massive loss of floral resource availability, and even early stages of forest succession can cause declines in floral abundance and diversity, with corresponding effects on bumble bees [@Iserbyt2015-jc; @Walcher2017-kw; @Suzuki-Ohno2020-jv]. In contrast, more accessible areas at the bases of mountains are undergoing agricultural intensification that displaces natural vegetation and favors grasses over forbs [@Tasser2002-om]. Caught between adverse land use changes above and below and subject to climate-induced range shifts, mountain bumble bees are facing a future of reduced floral resource availability and altered sympatry, and the ability of bumble species to coexist under these changes will likely depend significantly on how they select and partition floral resources.

In this study, we analyze a data set consisting of ~13,000 interactions between bumble bees and wildflowers recorded over 3 years along a 1400 m elevation gradient. Using these data in conjunction with complementary floral survey data, we explore the *selectivity* of the bumble bee community at the level of floral morphotype [@Kugler1970-gb].

**This is a good place to get our terminological ducks in a row**

We opt for the term *selectivity*, defined as the  (*sensu* @Beyer2010-vr) rather than the ... defined in this way, selectivity has the same conceptual meaning (though different mathematical derivation) as @Bluthgen2006-if *discrimination*.

We first compare the aggregate preferences of the bumble bee community across floral morphotypes and then model the preferences for each morphotype as a function of bumble bee tongue length. Then, by calculating preferences based on individual samples distributed through elevation and time, we model the variation of bumble bee preferences through temporal and elevational axes of floristic turnover, revealing the lability of bumble bee floral preferences. The balance of flexibility and constraint revealed in these analyses **Figure out how to finish this...**

Our investigation also affords an opportunity to revisit an open question regarding the role of resource abundance as a driver of flower selection in bumble bees. In studies of resource selection, it is common to pose an abundance-only hypothesis in which foraging is random and selection merely a function of frequency of encounter, resulting in resource use patterns that mirror resource abundance patterns. For bumble bees, which are known to forage non-randomly, the abundance-only hypothesis can serve as a null-model against which observed use patterns can be compared to infer *selectivity*, i.e. resource use that is either positively or negatively disproportionate to resource abundance. It has also be proposed, however --- in what has been termed the *target use hypothesis* --- that bumble bees might select resources irrespective of abundance [@De_Valpine2013-be; @Harmon-Threatt2017-bz]. This hypothesis, while obviously subject to boundary conditions of floral availability, is consistent with the observation that bumble bees sometimes persist in visiting favored flower species even when these resources are experimentally reduced to low abundance [@Heinrich1976-vb], and it was supported in a study comparing competing models of flower selection in *Bombus vosnesenskii* [@Harmon-Threatt2017-bz]. If and when the *target use hypothesis* holds, uncorrected visitation rates are a better indication of preference than the residuals of an abundance-only null model, since the subtraction of presumed neutral effects of abundance when those effects do not, in fact, exist actually creates new abundance-dependent artifacts and obscures the true process of resource selection [@Harmon-Threatt2017-bz]. On the other hand, failing to account for neutral effects of abundance when they do obtain similarly obscures the selection process underpinning foraging patterns [@Vazquez2003-qa; @Bluthgen2008-rp]. In our study, we take advantage of the fact that there exists are well-established mechanistic basis for bumble bee flower choice, namely the compatibility between floral morphology and bumble bee tongue length. Taking this mechanism as a reference, we evaluate models of preference with and without abundance effects by comparing the explanatory power of bumble bee tongue length with respect to the preferences inferred under each model.

*"[. . .] preference  is  the  ratio  of  the  use  of  a  habitat  over  its availability, conditional on the availability of all habitats to the study animals." [@Aarts2008-fo]*

One way we could handle our language and caveats is this. The kind of null model approach we used *does not* condition visitation to each focal resource on the abundances of all other resources, only on the proportional abundance of the focal resource. Following @Aarts2008-fo, we cannot then speak of *preference*. Instead, what we have is weaker concept of *selectivity* defined by Beyer2010-vr. We can, therefore, use the word *selectivity* after explaining its relation to preference in the strict sense. I do not want to go back and fit some crazy hierarchical model that conditions on the abundance of all morphotypes.

# 2. METHODS

## 2.1 Field sampling

Field sampling was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures in a coniferous forest matrix. We selected 25 study sites (60 x 60 m) on pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management.

Sampling consisted of repeated visits (hereafter, “visitation samples”) to each study site at approximately weekly intervals. Sampling was only conducted during periods without rain and with an air temperature of at least 6°C. Bumble bees were recorded during one 50-minute transect walk per visit, and bumblebees observed on a given flower were counted as floral visitors. Queens were identified to species level in the field to avoid destructively sampling colony foundresses, but workers and males were collected and frozen for later identification in the laboratory. Species were identified following @Amiet1996-jz. While male bumble bees was recorded, we chose to analyze only the pooled visitation of by queens and workers. In conjunction with bumble bee observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^ and identified the plant species according to @Lauber2007-ha and @Oberndorfer2001-ro.

While per-sample effort was standardized across all years, the total number of samples varied: 153 in 2010, 212 in 2011, and 539 in 2012. The start of field sampling each year was determined by snowmelt, so lower sites were sampled as early as April while higher sites could not be sampled until June or July. In 2010 and 2012, this resulted in a temporally staggered pattern of sampling, but a combination of early snowmelt and a late start to field activities in 2011 resulted in sampling that was approximately aligned in time across the whole elevation gradient. Due to these annual differences in sampling intensity and duration, we opted to pool data across years rather than include year effects in our models, which would multiply model complexity and be difficult to interpret. This means, however, that our analysis cannot detect year-to-year variation in floral preferences and niche overlap, and main effects cannot be distinguished from annual anomalies. With only three years of data, though, such annual inferences would be of little value, and we deemed them not worth the cost of added model complexity.

The samples analyzed in this paper have been previously analyzed (Sponsler et al., submitted), but the analyses presented here are non-overlapping.

## 2.2 External data sources

### 2.2.1 Floral morphology

Rather than considering various floral morphological traits independently, we relied on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Farnberger1996-xo; @Neumayer1999-il; @Schneller2014-ry].

Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database [@Klotz2002-uc]. Morphotypes were simplified to the primary classes numbered 0-10: 0 = null (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. In the case of the genus *Phyteuma*, we followed @Neumayer1999-il and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and morphotype 8 was completely absent, so these morphotypes were omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, and they were rarely visited by bumble bees, so this morphotype was also omitted from all analyses.

### 2.2.2 Bumble bee tongue length

Bumble bee tongue-length data (for the worker caste) were obtained from the literature values compiled and summarized by @Arbetman2017-xj. Values for *B. gerstaeckeri* and *B. mendax*, which were not included in @Arbetman2017-xj, were obtained from @Obeso1992-yp and @Durieux2000-uk, respectively.

## 2.3 Data processing

Prior to analysis, we aligned each visitation sample with a corresponding floral survey sample. In most cases, floral surveying was conducted jointly with bumble bee observation for each site and date. Occasionally, though, floral surveying could not be completed on the same day as bumble bee observation, resulting in a small temporal offset between visitation samples and floral survey samples. In these cases, visitation samples were paired with the nearest available floral survey data within a maximum distance of 7 days. Visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis, which resulted in dropping 40 site-dates out of a total of 895, leaving 855 site-dates included in our analyses. In a few cases, two consecutive visitation samples at a given site shared the same nearest floral survey sample; in these cases, the visitation samples were first pooled, then aligned to their shared nearest floral survey sample. In rare cases, a given floral species was recorded during visitation sampling but missed during floral surveying, and these species were retroactively assigned the minimum floral area value of 0.01 m^2^.

Having thus aligned visitation data and flower cover data, we summed visitation and floral cover across species within each floral morphotype. This yielded, for each of our 855 aligned samples, a cumulative value of flower cover for each floral morphotype and a cumulative value of visitation for each bumble- bee-morphotype pair.

## 2.4 Analysis of resource selection

**Start this by working out my terminology. Are we saying preference or bias? And what about the distinction between preference, use, and selection? See @Johnson1980-co and @Beyer2010-vr as authorities on terminology. Whatever I choose, I then need to go revise the labels of the figures accordingly. I think we could make an argument for sticking wit the word "bias" to describe the inferences of our null model analysis, since we are allowing that the raw visitation patterns may be as good or better a measure of "preference". We don't actually find that, of course, but I think we should avoid calling the null model inferences "preference" as though it were a foregone conclusion that they represent preference while the uncorrected visitation rates do not. But preference, alas, is the word the litertaure seems to have settled on, so when I discuss my conclusions, I should talk about preference, comparing preference as inferred from bias and preference as inferred from uncorrected visitation rates.**

*We will use the word "preference" in the phenomenological sense of @Beyer2010-vr to denote a disproportionality between resource availability and resource use while remaining agnostic as to the behavioral basis of this pattern. Or perhaps we will use the word bias but follow the same definition, since "bias" does not bear the teleological connotations of "preference". Double check that this is really how @Beyer2010-vr define preference. If so, it differs a bit from the more original definition of .*

A key challenge when interpreting interaction networks is to disentangle directed processes, such as a focal species' preference or avoidance of certain interaction partners, from the neutral effects of relative abundance [@Vazquez2003-qa; @Bluthgen2008-rp]. For example, in the context of a bumble bee floral visitation network such as ours, abundant floral taxa should be expected to receive more visitation than rare taxa simply because they are more frequently encountered, and it would clearly be a mistake to interpret this as meaningful "preference" or "specialization" on the part of the bumble bees. One way to control for the neutral effects of relative abundance is to compare observed visitation patterns to the simulated visitation patterns of a null model in which plants are visited randomly in proportion to their relative abundance [@Vazquez2003-qa; @Bluthgen2008-rp]. When constructing null models of interaction networks that are distributed through space and/or time, though, it is critical to respect the sampling unit of individual sites and dates, since only within this site-date sampling unit does floral visitation data correspond unequivocally to floral abundance data. Naive aggregation across sites and dates introduces forbidden links and decouples visitation and abundance [@Vaughan2018-gj].

We constructed null models following @Vaughan2018-gj. First, resource selection is simulated for individual consumers --- in our case, individual bumble bees --- by according to a resource selection process in which the probability of a resource being chosen is proportional to its relative abundance [@Agusti2003-tt]. By simulating the visitation of individual consumers rather than species in aggregate, the abundance of each resource, and therefore its probability of visitation, can vary across site-dates, thus preserving the within-sample correspondence between visitation and abundance. Simulated interactions are then summarized for each consumer species by taking the mean number of interactions with each resource across all individuals of the consumer species and all samples of the dataset. As this process is iterated, the result is a distribution of visitation rates of each consumer species to each resource under the null assumption that resource selection is determined only by resource abundance. The visitation distributions of the null model can then be compared to observed visitation rates to infer positive or negative bias of each consumer toward each resource. Since the mean of the null distribution can be interpreted as the expected value under null assumptions, the difference between the mean null value and the observed value can be interpreted as a measure of positive or negative bias [@Vaughan2018-gj]. The statistical significance of this bias can be estimated as the proportion of simulated null values more extreme than the observed value.

We performed two rounds of null model analysis targeting different ecological questions. First, we followed the procedure described above, simulating our entire dataset in one modeling process to infer aggregate pattern of bias between bumble bee species and floral morphotypes. Then, we repeated the null model analysis for each site-date individually so that the bias between each bumble bee species and floral morphotype could be modeled as a variable through elevation and time. For both analyses, we used 500 iterations to achieve a robust sampling distribution, and we considered observed visitation rates to indicate significant bias if they fell outside the interval containing 99% of simulated values.

Using the aggregate null model results, we modeled the biases toward each floral morphotype as a function of bumble bee tongue length using hierarchical generalized additive modeling (HGAM) with floral morphotype as a "by" smoothing factor [@Pedersen2019-kf]. We then compared morphotype biases across bumble bee species by performing a principal components analysis and mapped bumble bee tongue length onto the ordination space by fitting a 2-dimensional GAM surface to the first two principal components.

With the sample-wise null model results, we modeled the variation of bias between bumble bee species and floral morphotypes through elevation and time. For this analysis, we included only the six most abundant and widespread bumble bee species in our study system (*B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii*) to ensure sufficient data points through elevation and time, and we modeled their biases only toward the 4 most visited floral morphotypes (bell, lip, flag, and head flowers). Biases for each bumble- bee-morphotype pair were modeled together in a single HGAM with a tensor-product interaction smooth between elevation and time and bumble- bee-morphotype pair as a "by" smoothing factor.

## 2.5 Software

All data processing and analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. Null models were constructed using `econullnetr` [@Vaughan2018-gj]. GAM analyses were performed with `mgcv` [@Wood_2017] and visualized with `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta2021-aa]. PCA was performed with the `vegan` [@Oksanen_2019] and visualized using `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Annotated R code is available in the Appendix X.

# 3. RESULTS
